Template for indicator documentation

### 2023-03-24 breareal.Rmd
Add changes to date format for consistency

# (PART\*) INDICATORS FROM THE ALPINE ASSESSMENT {.unnumbered}

@@

```{r, include=FALSE, message=FALSE}
library(knitr)
library(sf)
library(terra)
library(raster) # only needed for creating boundary boxes, which can be done using sf
library(dplyr)
library(lwgeom)
library(tmap)
library(dplyr)
#library(ggplot2)
#library(data.table)
knitr::opts_chunk$set(echo = TRUE)
```

*Author and date:* Anders Kolstad

```{r, echo=F}
Sys.Date()
```

@@

<br /> <br />

<hr />

## Introduction

This indicator describes the relative change in glacial extent for each of fire regions in Norway. A [slightly different indicator](https://ninanor.github.io/IBECA/faktaark.html#Areal_av_isbreer) was used in the alpine assessment.

## About the underlying data

The indicator uses the maps of glacier extent ( *breatlas* ) from [2018-2019](https://www.nve.no/vann-og-vassdrag/vannets-kretsloep/bre/publikasjoner-publications/breatlas-glacier-inventories/) and compares it to a similar map from 1947-1985. Both maps are owned by NVE (Norsk vassdrags og energidirektorat). The new map is based on Sentinel images, and the old map is a digitised version of the N50 maps, with some additional areal photo interpretation on top.

### Representativity in time and space

The data set (map) covers the entire Norwegian mainland. There are only two time steps, but we expect changes in glacial extent to be relatively linear or consistent over time. The older map may have underestimated the glacial extent due to a better chance of smaller snow and ice patches being captured by Sentinel contra aerial photo interpretations.

### Original units

Area units (e.g. km2)

### Temporal coverage

1947-2019

### Aditional comments about the dataset

None

## Ecosystem characteristic

### Norwegian standard

The indicator was initially assigned the *Abiotiske forhold* characteristic, but now it is considered under *Landskapsøkologiske mønstre*. The initial justificaton was that glacial melting would influence water availability downstream. Now we put more weight on glaciers as not just resevoirs of ice and water, but as a habitat, and the presence or abundance of a specific habitat can be though of as an aspect of landscape state.

### SEEA ECT

The indicator is similarly assiged to the SEEA ECT class *C1 Landscape characteristics*.

## Collinearities with other indicators

None that are known.

## Reference state and values

### Reference state

The indicator uses the same reference state as all or most of the indicators developed in this and in similar projects, and is defined in [Nybø et al (2017): Fagsystem for fastsetting av god økologisk tilstand](https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/). As glacial extent is primarily, or solely, dependent on climate, we are in a case where we have actual data from the climatic reference period which is the previous normal period 1961-1990.

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values

The reference value is zero change in glacial extent over time. The threshold for defining good ecological condition as set at a 40% reduction in glacial extent, as a default solution. This normalization follows a linear model, although a sigmoidal weibull function might be more appropriate.

@@

## Uncertainties

Here we add a 3% error on each side of the estimates, same as what is reported for the glacier maps from NVE.

## References

-   Andreassen, L.M., & Winsvold, S.H. (eds.). 2012. Inventory of Norwegian glaciers. NVE Rapport 38, Norges Vassdrags- og energidirektorat, 236 s

-   Andreassen, L.M., Elvehøy, H., Kjøllmoen, B. & Belart, J.M.C. 2020. Glacier change in Norway since the 1960s -- an overview of mass balance, area, length and surface elevation changes. Journal of Glaciology 66: 313--328

-   Andreassen, L.M., Nagy, T., Kjøllmoen, B. & Leigh, J.R. 2022. A Sentinel-2 based inventory of Norway's glaciers and ice-marginal lakes 2018/2019. Journal of Glaciology (in review)

-   Winsvold, S.H., Andreassen, L.M. & Kienholz, C.. 2014. Glacier area and length changes in Norway from repeat inventories. The Cryosphere 8: 1885-1903

@@

### Importing the data sets

There are four data sets:

-   glacial atlas
-   n50 (the old glacial atlas)
-   the regional delineation for Norway (fire regions)
-   an outline of mainland Norway

@@

#### Glacial atlas

This is the new glacial atlas from 2018 and 2019 based on Sentinell (ref. Liss Marie Andreassen, NVE)

@@

There are 6915 polygons. Let's calculate the area of the polygons and study the distribution of size classes.

```{r}
breatlas$area <- st_area(breatlas)
par(mfrow=c(1,2))
hist(breatlas$area)
plot(breatlas$area)
```

@@

#### Map of Norway

This map is stored on the GitHub repository under *indicators/data/*.

```{r}
nor <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")

@@

#### N50

This is the old glacial extent map. Import it and transform it, same as before.

@@

The red areas in the map to the far right is where the ice has retreated for a selected glacier on the west coast. Now we need to devide the map into regions to then compare the area in n50 with the new glacial extent map *breatlas*.

#### Regions

Importing a shape file with the regional delineation.

@@

There is a convinient function to fix this.

```{r}
# requires lwgeom
breatlas2 <- st_make_valid(breatlas)
table(st_is_valid(breatlas2))

@@

Let me plot the new, fixed map on top of the old to see if they match. I give the new valid map a blue transparent colour, and the old map a yellow and transparent colour.

```{r}
plot(nor$geometry, xlim=c(5000, 10000),
          ylim=c(6880000, 6885000),
     axes=T)
    plot(breatlas2$geometry, add=T, col = scales::alpha("blue",0.5))
    plot(breatlas$geometry, add=T, col = scales::alpha("yellow",0.5))
```

@@

Delete the invalid data set.

```{r}
breatlas <- breatlas2
rm(breatlas2)
```

### Glacial area per region

#### Contemporary area

Here is one method for getting the glacial area for each region.

```{r}
# This intersection operation is a bit heavy, so I do it once and store it locally so that I can import it again quickly. 
#brealtas_reg <- st_intersection(breatlas, reg)
#saveRDS(brealtas_reg, "indicators/data/brealtas_reg_helperfile.rds")

@@

Checking what happens to polygons that span to regions. Central Norway in red and western Norway in green. 50% transparency means we should get a third colour where and if they overlap.

```{r}
plot(brealtas_reg$geometry[brealtas_reg$region=="Midt-Norge"], 
    col = scales::alpha("red",0.5), border=NA, axes=T, 
     ylim=c(6900000, 6905000),
     xlim=c(85000, 100000))
plot(brealtas_reg$geometry[brealtas_reg$region=="Vestlandet"], 
     col = scales::alpha("green",0.5), border=NA, add=T)
  plot(nor$geometry, add=T)
```

@@

Calculating the area of the polygons

```{r}
#Calculating the area of the polygons
brealtas_reg$area_crop <- st_area(brealtas_reg)

@@

Southern Norway has \<1km2 of glaciers in 2018-2019.

#### Same for N50

```{r}
n50x <- st_make_valid(n50)
#n50_reg <- st_intersection(n50x, reg)
#saveRDS(n50_reg, "indicators/data/n50_helperfile.rds")
n50_reg <- readRDS("data/n50_helperfile.rds")
```

@@

Thats good. Now I will plot the reference values (n50) and the contemporary data together.

```{r}
n50_reg$area_crop <- st_area(n50_reg)

@@

### Scaled indicator values

Making a table with the indicator values.

```{r}
myTbl <- tibble(region           = names(bretabell),
                indikator        = bretabell,
                referanseverdi   = n50tabell)
myTbl$skalert_indikator <- myTbl$indikator/myTbl$referanseverdi

@@

Plotting a static map with scaled indicator values

```{r}
tmap_mode("plot") # mode 'view' takes a lot longer to render, but looks better
tm_shape(reg_clipped) + 
  tm_polygons(col="skalert_indikator", 
              border.col = "white")+
  tm_text("skalert_indikator_r", size=1.5)+
  tm_shape(nor)+
  tm_polygons(alpha = 0,border.col = "black")
```

@@

Calculating total change of glacier extent

```{r}
sum(bretabell)/sum(n50tabell)
```

@@

### Uncertainty

NVE reports a 3% uncertainty with their method. This is interpret as if +- 3% represent a 99.9%CI for the estimate (ca 2 SD). In the [original indicator](https://ninanor.github.io/IBECA/areal-av-isbreer.html) I re-sampled the indicator value 10000 times here, with a 0.015sd, to get a distribution of indicator values. This time I will settle for reporting the sd as a single value below.

## Prepare export

Now I can prepare the final exported data file. It will be a shape file with raw and scaled indicator values, reference reference values, and uncertainties.

>>>>

# (PART\*) INDICATORS FROM THE ALPINE ASSESSMENT {-}

@@

```{r, include=FALSE, message=FALSE}
library(knitr)
library(sf)
library(terra)
library(raster) # only needed for creating boundary boxes, which can be done using sf
library(dplyr)
library(lwgeom)
library(tmap)
library(dplyr)
#library(ggplot2)
#library(data.table)
knitr::opts_chunk$set(echo = TRUE)
```
<br />

_Author and date:_

Anders L. Kolstad

March 2023

@@

<br />
<br />
<hr />

## Introduction
This indicator describes the relative change in glacial extent for each of fire regions in Norway. A [slightly different indicator](https://ninanor.github.io/IBECA/faktaark.html#Areal_av_isbreer) was used in the alpine assessment.

## About the underlying data
The indicator uses the maps of glacier extent ( _breatlas_ ) from [2018-2019](https://www.nve.no/vann-og-vassdrag/vannets-kretsloep/bre/publikasjoner-publications/breatlas-glacier-inventories/) and compares it to a similar map from 1947-1985. Both maps are owned by NVE (Norsk vassdrags og energidirektorat). The new map is based on Sentinel images, and the old map is a digitised version of the N50 maps, with some additional areal photo interpretation on top.

### Representativity in time and space
The data set (map) covers the entire Norwegian mainland. There are only two time steps, but we expect changes in glacial extent to be relatively linear or consistent over time.
The older map may have underestimated the glacial extent due to a better chance of smaller snow and ice patches being captured by Sentinel contra aerial photo interpretations.

### Original units
Area units (e.g. km2)

### Temporal coverage
1947-2019

### Aditional comments about the dataset
None

## Ecosystem characteristic
### Norwegian standard
The indicator was initially assigned the _Abiotiske forhold_ characteristic, but now it is considered under _Landskapsøkologiske mønstre_. The initial justificaton was that glacial melting would influence water availability downstream. Now we put more weight on glaciers as not just resevoirs of ice and water, but as a habitat, and the presence or abundance of a specific habitat can be though of as an aspect of landscape state.

### SEEA ECT
The indicator is similarly assiged to the SEEA ECT class _C1 Landscape characteristics_.

## Collinearities with other indicators
None that are known.

## Reference state and values
### Reference state
The indicator uses the same reference state as all or most of the indicators developed in this and in similar projects, and is defined in [Nybø et al (2017): Fagsystem for fastsetting av god økologisk tilstand](https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/). As glacial extent is primarily, or solely, dependent on climate, we are in a case where we have actual data from the climatic reference period which is the previous normal period 1961-1990.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
The reference value is zero change in glacial extent over time. The threshold for defining good ecological condition as set at a 40% reduction in glacial extent, as a default solution. This normalization follows a linear model, although a sigmoidal weibull function might be more appropriate.

@@

## Uncertainties
Here we add a 3% error on each side of the estimates, same as what is reported for the glacier maps from NVE.

## References
* Andreassen, L.M., & Winsvold, S.H. (eds.). 2012. Inventory of Norwegian glaciers. NVE Rapport 38, Norges Vassdrags- og energidirektorat, 236 s

* Andreassen, L.M., Elvehøy, H., Kjøllmoen, B. & Belart, J.M.C. 2020. Glacier change in Norway since the 1960s – an overview of mass balance, area, length and surface elevation changes. Journal of Glaciology 66: 313–328

* Andreassen, L.M., Nagy, T., Kjøllmoen, B. & Leigh, J.R. 2022. A Sentinel-2 based inventory of Norway’s glaciers and ice-marginal lakes 2018/2019. Journal of Glaciology (in review)

* Winsvold, S.H., Andreassen, L.M. & Kienholz, C.. 2014. Glacier area and length changes in Norway from repeat inventories. The Cryosphere 8: 1885-1903

@@

### Importing the data sets
There are four data sets:

* glacial atlas
* n50 (the old glacial atlas)
* the regional delineation for Norway (fire regions)
* an outline of mainland Norway

@@

#### Glacial atlas
This is the new glacial atlas from 2018 and 2019 based on Sentinell (ref. Liss Marie Andreassen, NVE)
```{r}

@@

There are 6915 polygons. Let's calculate the area of the polygons and study the distribution of size classes. 
```{r}
breatlas$area <- st_area(breatlas)
par(mfrow=c(1,2))
hist(breatlas$area)
plot(breatlas$area)
```

@@

#### Map of Norway
This map is stored on the GitHub repository under _indicators/data/_. 
```{r}
nor <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")

@@

#### N50
This is the old glacial extent map. Import it and transform it, same as before.

@@

The red areas in the map to the far right is where the ice has retreated for a selected glacier on the west coast. 
Now we need to devide the map into regions to then compare the area in n50 with the new glacial extent map _breatlas_.

#### Regions
Importing a shape file with the regional delineation.

@@

There is a convinient function to fix this.
```{r}
# requires lwgeom
breatlas2 <- st_make_valid(breatlas)
table(st_is_valid(breatlas2))

@@

Let me plot the new, fixed map on top of the old to see if they match. I give the new valid map a blue transparent colour, and the old map a yellow and transparent colour.  
```{r}
plot(nor$geometry, xlim=c(5000, 10000),
          ylim=c(6880000, 6885000),
     axes=T)
    plot(breatlas2$geometry, add=T, col = scales::alpha("blue",0.5))
    plot(breatlas$geometry, add=T, col = scales::alpha("yellow",0.5))
```

@@

Delete the invalid data set.
```{r}
breatlas <- breatlas2
rm(breatlas2)
```

### Glacial area per region
#### Contemporary area

Here is one method for getting the glacial area for each region.
```{r}
# This intersection operation is a bit heavy, so I do it once and store it locally so that I can import it again quickly. 
#brealtas_reg <- st_intersection(breatlas, reg)
#saveRDS(brealtas_reg, "indicators/data/brealtas_reg_helperfile.rds")

@@

Checking what happens to polygons that span to regions. Central Norway in red and western Norway in green. 50% transparency means we should get a third colour where and if they overlap.
```{r}
plot(brealtas_reg$geometry[brealtas_reg$region=="Midt-Norge"], 
    col = scales::alpha("red",0.5), border=NA, axes=T, 
     ylim=c(6900000, 6905000),
     xlim=c(85000, 100000))
plot(brealtas_reg$geometry[brealtas_reg$region=="Vestlandet"], 
     col = scales::alpha("green",0.5), border=NA, add=T)
  plot(nor$geometry, add=T)
```

@@

Calculating the area of the polygons
```{r}
#Calculating the area of the polygons
brealtas_reg$area_crop <- st_area(brealtas_reg)

@@

Southern Norway has <1km2 of glaciers in 2018-2019.

#### Same for N50
```{r}
n50x <- st_make_valid(n50)
#n50_reg <- st_intersection(n50x, reg)
#saveRDS(n50_reg, "indicators/data/n50_helperfile.rds")
n50_reg <- readRDS("data/n50_helperfile.rds")
```

@@

Thats good. 
Now I will plot the reference values (n50) and the contemporary data together.
```{r}
n50_reg$area_crop <- st_area(n50_reg)

@@

### Scaled indicator values
Making a table with the indicator values.
```{r}
myTbl <- tibble(region           = names(bretabell),
                indikator        = bretabell,
                referanseverdi   = n50tabell)
myTbl$skalert_indikator <- myTbl$indikator/myTbl$referanseverdi

@@

Plotting a static map with scaled indicator values
```{r}
tmap_mode("plot") # mode 'view' takes a lot longer to render, but looks better
tm_shape(reg_clipped) + 
  tm_polygons(col="skalert_indikator", 
              border.col = "white")+
  tm_text("skalert_indikator_r", size=1.5)+
  tm_shape(nor)+
  tm_polygons(alpha = 0,border.col = "black")
```

@@

Calculating total change of glacier extent
```{r}
sum(bretabell)/sum(n50tabell)
```

@@

### Uncertainty
NVE reports a 3% uncertainty with their method. This is interpret as if +- 3% represent a 99.9%CI for the estimate (ca 2 SD).
In the [original indicator](https://ninanor.github.io/IBECA/areal-av-isbreer.html) I re-sampled the indicator value 10000 times here, with a 0.015sd, to get a distribution of indicator values. This time I will settle for reporting the sd as a single value below.

## Prepare export
Now I can prepare the final exported data file. It will be a shape file with raw and scaled indicator values, reference reference values, and uncertainties.
